Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2025-02-25 analysis/paper/paper.qmd
add figure captions and sentences

Surface modifications, including cut marks, fractures, and burning damage, were identified on deer bones from CL6 to CL2, with variations in their proportions (@tbl-deer-modify). Burned bones, characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4. Cut marks were present on 50% of bones in CL2, 30% of bones in earlier period (CL6 to CL4), and at a lower proportion of 23% in CL3. Most cut marks were located on major limb joints and long bones, suggesting butchering activities related to disarticulation and meat detaching. @fig-deer-cutmark shows that at least 50 % of joints of each deer part in CL4 present cut marks, with a higher proportion of 75% at the hip. A similar pattern was observed in CL5, but with a lower proportion of cut marks on the ankle, knee, and hip compared to CL4. Not all joints were found in CL6, CL3 and CL2, making direct comparisons across all layers difficult. Nevertheless, the overall high proprotion of joints with cut marks suggests intensive proceessing of deer throughout the site's occupation. Cut marks on lower limbs can be associeted with skinning and were primarily found on metapodials in CL6 (n = 1), CL5 (n = 4), CL4 (n = 6), CL3 (n = 1), followed by first phalanges in CL4 (n = 1) and CL5 (n = 1).

Bone breakage patterns, including spiral fractures, transverse fractures, longitudinal splits, percurssion scars, and chop marks [@Haynes1983; @Gifford2018; @Fisher1995], were observed esepically on long bones across different layers. These fractures indicate intentional marrow extraction at Kiwulan. CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5. This suggests more intensive marrow extraction in later periods. Most chop and cut marks were likly made by metal tools according to clean, smooth cut surfaces and finer marks on bones. The use of metal tools can be further supported by the finding of iron knives at Kiwulan [@Chen2007]. In addition to anthropogenic modifications, natural alterations to bones resulting from post-deposition impacts, such as weathering and root etching [@Gifford2018natural], were identified on approximately one-third of the bones across all layers (@tbl-deer-modify).

@@

>>>>

Surface modifications, including cut marks, burning damage, and fractures were identified on deer bones from CL6 to CL2, with variations in their proportions (@tbl-deer-modify). Cut marks were present on 50% of bones in CL2, 30% of bones in earlier period (CL6 to CL4), and at a lower proportion of 23% in CL3. Most cut marks were located on major limb joints and long bones, suggesting butchering activities related to disarticulation and meat detaching. Some joints displyed chop marks along with cut marks (@fig-deer-cutmark), of which the clean, smooth cut surfaces and sharp marks suggest they were likly made by metal tools [@Greenfield1999]. @fig-deer-cutmark shows that at least 50 % of joints of each deer part in CL4 present cut marks, with a higher proportion of 75% at the hip. A similar pattern was observed in CL5, but with a lower proportion of cut marks on the ankle, knee, and hip compared to CL4. Not all types of joints were found in CL6, CL3 and CL2, making direct comparisons across all layers difficult. Nevertheless, the overall high proprotion of joints with cut marks suggests intensive proceessing of deer throughout the site's occupation. In addition, cut marks on lower limbs can be associeted with skinning (@fig-deer-cutmark: e) and were primarily found on metapodials in CL6 (n = 1), CL5 (n = 4), CL4 (n = 6), CL3 (n = 1), followed by first phalanges in CL4 (n = 1) and CL5 (n = 1).

Burned bones, characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4. Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percurssion scars [@Haynes1983; @Gifford2018; @Fisher1995], were observed on long bones across different layers (@fig-deer-fracture). These fractures indicate intentional marrow extraction at Kiwulan. CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5. This suggests more intensive marrow extraction in later periods. Chewing marks were also observed on some bones that show a series of shallow and smotth features (@fig-deer-fracture: e). In addition to anthropogenic modifications, natural alterations to bones resulting from post-deposition impacts, such as weathering and root etching [@Gifford2018natural], were identified on approximately one-third of the bones across all layers (@fig-deer-fracture: c,d; @tbl-deer-modify).

@@

```{r}
#| label: fig-cutmark-img
#| echo: false
#| fig-cap: "Examples of sika deer bones with cut marks, chop surfaces, and burning damage. Cut marks are indicated with red squares, and chop surfaces are indicated with black arrows. Bone fragments are (a) proximal radius with cut marks and a chop surface, (b) distal humerus with cut marks, (c) acetabulum with chop surfaces, (d) charred astragulus, (e) metatarsal with cut marks."

knitr::include_graphics(here::here("analysis/figures/deer_cutmark_img.png"))
```

```{r}
#| label: fig-fracture-img
#| echo: false
#| fig-cap: "Examples of sika deer bones with fractures, natural modifications, and chewing marks. Bone fragments are (a) proximal metatarsal with a puncturing hole and longitudinal split, (b) distal tibia showing spiral fracture, (c) distal radius with root etching, (d) femur shaft showing transverse fracture on the bottom, a percurssion scar in the middle with longitudinal split, and weathering, (e) scapula with chewing marks."

knitr::include_graphics(here::here("analysis/figures/deer_fracture_img.png"))
```
